Human Endogenous Retroviral Transcripts Show Potential as Novel, Tumor Specific Biomarkers for Colorectal Cancer

### 2021-01-25 manuscript/secs/discussion.tex
add cover letter, tweak discussion

In this study, we performed the largest unbiased assessment of transcribed, unannotated human endogenous retroviral elements in colorectal tissue to date.
We found several unique proviral loci with tumor specific over-expression and other loci with relatively high expression in healthy samples.
Human endogenous retroviral elements are often transcriptionally repressed repetitive sequences that are frequently de-repressed in epigenetically altered cancer cells \citep{Babaian2016}.
HERV transcripts have been shown to engage pattern recognition receptors and activate interferon responses in cancer cell lines \citep{Chiappinelli2015, Roulois2015} and have been associated with tumor infiltration by T regulatory cells \citep{Desai2017}, response to immune checkpoint blockade in urothelial carcinoma, and survival in subjects with CRC \citep{Solovyov2018}.
Our results are important because the tumor specific HERV sequences we identified could be used to distinguish tumor from non-tumor tissue in screening and surveillance and could be used to predict or even modify response to immunotherapies.

Three of the seven HERV proviruses we validated as tumor specific were from group H.
Over expression of HERV-H proviruses in CRC has been demonstrated previously with microarray \citep{Perot2012} and quantitative real time polymerase chain reaction \citep{Perot2015}.
The prominence of HERV-H transcripts in RNA-seq from TCGA COAD samples has also been previously reported \citep{Desai2017, Attig2019}.
Therefore, our findings are generally consistent with prior work.
However, we provide new evidence of the reliability of tumor specific over expression by jointly analyzing a large number of samples from different studies.
Furthermore, our analysis was not restricted to a particular group of proviruses.
The unbiased nature of our assessment provides important context for selecting genomic features to develop as potential biomarkers or immunotargets for CRC.

@@

This study also demonstrated widespread expression of a subset of HERVs in healthy colorectal tissue.
Interpretation of this result is challenging.
HERVs can have rare functional roles in healthy cells \citep{Rote2004} and could conceivably have a role in maintaining healthy colorectal epithelium, but this is speculative.
Over expression in healthy tissue may be a characteristic of the subset of HERV transcripts selected for analysis.
Selected transcripts were previously unannotated and understudied, so their patterns of expression are largely unknown.
Notwithstanding the biological implications of high levels of HERV expression in healthy samples, the bioinformatic implications were reassuring.
Transcripts of repeat elements included in this study were assembled exclusively from contigs of cancer tissue \citep{Attig2019}, which might be expected to favor the detection of widespread tumor specific over-expression.
The over-expression of HERVs in healthy tissue reduced the likelihood of this bias.

A major limitation of this study was its reliance on heterogeneous RNA-seq datasets and quantifications of short reads of repetitive sequences that are difficult to estimate accurately.
Expression differences may have resulted from technical or demographic differences between phenotypes.
An additional limitation is the prominence of TCGA samples in all three cohorts studied, which could compromise independence of expression signals across cohorts despite independence of subjects.
To address these limitations, we built a sophisticated bioinformatic pipeline based on state-of-the-art software.
Specifically, our definition of HERV elements was of critical importance.
We used a published, hidden Markov model based annotation of a \emph{de novo} assembled cancer transcriptome to define and locate the HERV transcripts and loci we studied \citep{Attig2019}.
We used a published, alignment-free, k-mer matching quantification algorithm to probabilistically estimate transcript abundance of repetitive sequences that may otherwise have been excluded from analysis due to multiple possible mappings \citep{Patro2017}.
We used latent factor analysis to effectively adjust for multiple confounders, including subject age \citep{sva}.
We used actively developed and updated statistical packages designed specifically for high dimensional genomics analyses to perform association tests \citep{Love2014, sva}.
We also pursued multiple validations of our findings and methods to ensure reproducibility.

>>>>

In this study, we performed the largest unbiased assessment of transcribed, unannotated human endogenous retroviral elements in colorectal tissue to date.
We found several unique proviral loci with tumor specific over-expression and other loci with relatively high expression in healthy samples.
Human endogenous retroviral elements are often selectively de-repressed in epigenetically altered cancer cells and may be useful as tumor markers \citep{Babaian2016}.
HERV transcripts have been shown to engage pattern recognition receptors and activate interferon responses in cancer cell lines \citep{Chiappinelli2015, Roulois2015}, and they have been negatively associated with tumor infiltration by regulatory T cells \citep{Desai2017}.
HERV transcripts have also been positively associated with response to immune checkpoint blockade for urothelial carcinoma and negatively associated with survival without immunotherapy for CRC \citep{Solovyov2018}.
Although a consistent relationship between HERV expression and cytotoxic activity was not apparent in our bulk RNA-seq data, we identified a subset of HERV transcripts that performed well as tumor markers and varied across MSI subtypes.
Our results are important because the tumor specific HERV sequences we identified could be used to distinguish tumor from non-tumor tissue in screening and surveillance and could be used to predict or even modify response to immunotherapies.

Three of the seven HERV proviruses we validated as tumor specific were from group H.
Over expression of HERV-H proviruses in CRC has been demonstrated previously with microarray \citep{Perot2012} and quantitative real time polymerase chain reaction \citep{Perot2015}.
The prominence of HERV-H transcripts in RNA-seq from TCGA COAD samples has also been previously reported \citep{Desai2017, Attig2019}.
Therefore, our findings are generally consistent with prior work.
However, we provide new evidence of the reliability of tumor specific over-expression by jointly analyzing a large number of samples from different studies.
Furthermore, our analysis was not restricted to a particular group of proviruses.
The unbiased nature of our assessment provides important context for selecting genomic features to develop as potential biomarkers or immunotargets for CRC.

@@

This study also demonstrated widespread expression of a subset of HERVs in healthy colorectal tissue.
HERVs can have rare functional roles in healthy cells \citep{Rote2004} and could conceivably have a role in maintaining healthy colorectal epithelium, but this is speculative.
Alternatively, HERV loci may undergo transcription in healthy tissue without functional consequence.
Notwithstanding the biological implications of HERV expression in healthy samples, the bioinformatic implications were reassuring.
Transcripts of repeat elements included in this study were assembled exclusively from contigs of cancer tissue \citep{Attig2019}, which might be expected to favor the detection of widespread tumor specific over-expression.
The over-expression of HERVs in healthy tissue demonstrated the limited extent of this potential bias.

A major limitation of this study was its reliance on heterogeneous RNA-seq datasets and quantifications of short reads of repetitive sequences that are difficult to estimate accurately.
Expression differences may have resulted from technical or demographic differences between phenotypes.
An additional limitation was the prominence of TCGA samples in all three cohorts studied, which could compromise independence of expression signals across cohorts despite independence of subjects.
To address these limitations, we built a sophisticated bioinformatic pipeline based on state-of-the-art software.
Specifically, our definition of HERV elements was of critical importance.
We used a published, hidden Markov model based annotation of a \emph{de novo} assembled cancer transcriptome to define and locate the HERV transcripts and loci we studied \citep{Attig2019}.
We used a published, alignment-free, k-mer matching quantification algorithm to probabilistically estimate transcript abundance of repetitive sequences that may otherwise have been excluded from analysis due to multiple possible mappings \citep{Patro2017}.
We used latent factor analysis to effectively adjust for multiple confounders, including subject age \citep{sva}.
We used actively developed and updated statistical packages designed specifically for high dimensional genomics analyses to perform association tests \citep{Love2014, sva}.
We also pursued multiple validations of our findings and methods to ensure reproducibility.

### 2021-01-25 manuscript/secs/results.tex
add cover letter, tweak discussion

First, to confirm the ability of the 11 HERVs to distinguish tumor from non-tumor tissue, we used their expression level to classify samples.
As expected, when tested on the same samples from which their tumor specificity was discovered, most of the HERVs had excellent performance as biomarkers (Supplementary Figure 2).
Second, to determine whether the age difference across phenotypes was likely to have artificially inflated the apparent tumor specificity of the markers, we measured the correlation of HERV expression level with subject age.
Although a weak positive correlation was apparent (Supplementary Figure 3a), comparison of test statistics from linear models predicting expression level from age and phenotype showed a negligible effect of age (Supplementary Figure 3b).
Third, to assess the relationship between HERV expression and tumor characteristics, we plotted HERV expression separately for each category of tumor microsatellite instability status and stage.
The tumor specific HERVs tended to be higher in microsatellite stable tumors with the notable exceptions of HERV-H Xp22.32 and HERV-H 20p11.23 (Supplementary Figure 4a).
No trend in expression level relative to tumor stage was apparent (Supplementary Figure 4b).
Finally, to determine the relationship between HERV expression and cytotoxic activity within tumors, we measured the correlation between HERV expression and the expression of six markers of cytotoxic activity.
The correlation was minimal for most HERVs, but HERV1-I 3p22.3 had a weakly negative association (Supplementary Figure 5).

>>>>

First, to confirm the ability of the 11 HERVs to distinguish tumor from non-tumor tissue, we used their expression level to classify samples.
As expected, when tested on the same samples from which their tumor specificity was discovered, most of the HERVs had excellent performance as biomarkers (Supplementary Figure 2).
Second, to determine whether the age difference across phenotypes was likely to have artificially inflated the apparent tumor specificity of the markers, we measured the correlation of HERV expression level with subject age.
Although a weak positive correlation was apparent (Supplementary Figure 3a), comparison of test statistics from linear models predicting expression level from age and phenotype showed a negligible effect of age (Supplementary Figure 3b).
Third, to assess the relationship between HERV expression and tumor characteristics, we plotted HERV expression separately for each category of tumor microsatellite instability (MSI) status and stage.
The tumor specific HERVs tended to be higher in microsatellite stable tumors with the notable exceptions of HERV-H Xp22.32 and HERV-H 20p11.23 (Supplementary Figure 4a).
No trend in expression level relative to tumor stage was apparent (Supplementary Figure 4b).
Finally, to determine the relationship between HERV expression and cytotoxic activity within tumors, we measured the correlation between HERV expression and the expression of six markers of cytotoxic activity.
The correlation was minimal for most HERVs, but HERV1-I 3p22.3 had a weakly negative association (Supplementary Figure 5).
